Categorical Foundations of E-Graphs With Binding

### 2025-07-09 POPL2026/paper/sections/combinatorial_semantics.tex
notes and popl template
comparison, semantics, enrichment fix, appendix
figures and benchmarks

\begin{figure}[t!]
    \begin{subfigure}[T]{0.35\linewidth}
    \[
                % \tikzfig{../figures/combinatorial_semantics/DPO_square}
                \begin{tikzcd}
                    {\mathcal{L}} & {\mathcal{K}} & {\mathcal{R}} \\
                    {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}}
                    \arrow["f"', from=1-1, to=2-1]
                    \arrow[from=1-2, to=1-1]
                    \arrow[from=1-2, to=1-3]
                    \arrow[from=1-2, to=2-2]
                    \arrow[from=1-3, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
                    \arrow[from=2-2, to=2-1]
                    \arrow[from=2-2, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
                \end{tikzcd}
    \]
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.4\linewidth}
        \[
    % \scalebox{0.8}{
        % \tikzfig{../figures/combinatorial_semantics/DPOI_square_Hyp}
    % }
    \begin{tikzcd}
        {\mathcal{L}} & {i+j} & {\mathcal{R}} \\
        {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}} \\
        & {n+m}
        \arrow["f"', from=1-1, to=2-1]
        \arrow[from=1-2, to=1-1]
        \arrow[from=1-2, to=1-3]
        \arrow[from=1-2, to=2-2]
        \arrow[from=1-3, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
        \arrow[from=2-2, to=2-1]
        \arrow[from=2-2, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
        \arrow[from=3-2, to=2-1]
        \arrow[from=3-2, to=2-2]
        \arrow[from=3-2, to=2-3]
    \end{tikzcd}
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \hfill
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

@@

\section{Closed E-hypergraphs}
Prior work of~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established body of literature on hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The first work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

@@

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{e-color}{<}, \textcolor{closed-color}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$ is a set of hyperedges that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{e-color}{<} \subseteq \textcolor{e-color}{E} \times V + E$ and $\textcolor{closed-color}{<} \subseteq \textcolor{closed-color}{E} \times V + E $ with $\textcolor{e-color}{E}, \textcolor{closed-color} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{e-color}{<^{\mu}}$ and $\textcolor{closed-color}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{e-color}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{closed-color}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{e-color}{<^{\mu}} x$ or $e \textcolor{e-color}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{e-color}{<^{\mu}} e$ iff $e' \textcolor{e-color}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{closed-color}{<}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{e-color}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{e-color}{<^{\mu}}(v)$ and $p \textcolor{e-color}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{e-color}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{e-color}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{e-color}{E}$ and $\textcolor{closed-color}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{e-color}{<}, \textcolor{closed-color}{<}$ and~$\consistency$.
\end{definition}

@@

Consider an example in the middle part of Fig.~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{e-color}{E} = \{e_{1}\}$, $\textcolor{closed-color}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\vspace{-1ex}
\begin{array}{cccc}
  v_{4} \textcolor{e-color}{<^{\mu}} e_{1} & v_{7} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{e-color}{<^{\mu}} e_{1} & v_{8} \textcolor{closed-color}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{e-color}{<^{\mu}} e_{1} & e_{5} \textcolor{closed-color}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{e-color}{<^{\mu}} e_{1} & w_{9} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]
and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{e-color}{E}$ are depicted with a dashed box, while edges in $\textcolor{closed-color}{E}$ --- with a round solid box.
\begin{figure}
\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2}
}
\]
\captionsetup{skip=0pt, belowskip=-1ex}
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}
\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{e-color}{E_{\mathcal{F}}}) \subseteq \textcolor{e-color}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{closed-color}{E_{\mathcal{F}}}) \subseteq \textcolor{closed-color}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          \text{if } e \textcolor{e-color}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{e-color}{<_{\mathcal{G}}^{\mu}} \phi(v)
          \qquad
          \text{if } e \textcolor{closed-color}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          \text{if } e_1 \textcolor{e-color}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{e-color}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \qquad
          \text{if } e_1 \textcolor{closed-color}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$.
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

@@

\end{definition}
We call $n$  \textit{external input interfaces}, $n'$ \textit{internal input interfaces},
and $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  
We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n$ and $m$ with their images in $n'$ and $m'$,  and likewise $m, m, n', m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.
Graphically, we depict such extended cospans as in Fig.~\ref{fig:e-cospan-example}: feet of the cospan are depicted within shaded regions.
All non-black vertices form strictly internal interfaces.

>>>>

\begin{figure}[t!]
    \begin{subfigure}[T]{0.35\linewidth}
    \[
                % \tikzfig{../figures/combinatorial_semantics/DPO_square}
                \begin{tikzcd}
                    {\mathcal{L}} & {\mathcal{K}} & {\mathcal{R}} \\
                    {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}}
                    \arrow["f"', from=1-1, to=2-1]
                    \arrow[from=1-2, to=1-1]
                    \arrow[from=1-2, to=1-3]
                    \arrow[from=1-2, to=2-2]
                    \arrow[from=1-3, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
                    \arrow[from=2-2, to=2-1]
                    \arrow[from=2-2, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
                \end{tikzcd}
    \]
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.4\linewidth}
        \[
    % \scalebox{0.8}{
        % \tikzfig{../figures/combinatorial_semantics/DPOI_square_Hyp}
    % }
    \begin{tikzcd}
        {\mathcal{L}} & {i+j} & {\mathcal{R}} \\
        {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}} \\
        & {n+m}
        \arrow["f"', from=1-1, to=2-1]
        \arrow[from=1-2, to=1-1]
        \arrow[from=1-2, to=1-3]
        \arrow[from=1-2, to=2-2]
        \arrow[from=1-3, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
        \arrow[from=2-2, to=2-1]
        \arrow[from=2-2, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
        \arrow[from=3-2, to=2-1]
        \arrow[from=3-2, to=2-2]
        \arrow[from=3-2, to=2-3]
    \end{tikzcd}
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

@@

\section{Closed E-hypergraphs}
\label{sec:closed-e-hyp}
Prior work of~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established body of literature on hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The first work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

@@

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{e-color}{<}, \textcolor{closed-color}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$ is a set of hyperedges that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{e-color}{<} \subseteq \textcolor{e-color}{E} \times (V + E)$ and $\textcolor{closed-color}{<} \subseteq \textcolor{closed-color}{E} \times (V + E)$ with $\textcolor{e-color}{E}, \textcolor{closed-color}{E} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{e-color}{<^{\mu}}$ and $\textcolor{closed-color}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{e-color}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{closed-color}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{e-color}{<^{\mu}} x$ or $e \textcolor{closed-color}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{e-color}{<^{\mu}} e$ iff $e' \textcolor{e-color}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{closed-color}{<^{\mu}}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{e-color}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{e-color}{<^{\mu}}(v)$ and $p \textcolor{e-color}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{e-color}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{e-color}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{e-color}{E}$ and $\textcolor{closed-color}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$.
When writing $x \textcolor{closed-color}{<} y$ (or $x \textcolor{e-color}{<} y$) we may refer to $x$ as a predecessor of $y$ and refer to $y$ as a successor of $x$.
% we do not use this
% Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{e-color}{<}, \textcolor{closed-color}{<}$ and~$\consistency$.
\end{definition}

@@

Consider an example in the middle part of Fig.~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{e-color}{E} = \{e_{1}\}$, $\textcolor{closed-color}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\begin{array}{cccc}
  v_{4} \textcolor{e-color}{<^{\mu}} e_{1} & v_{7} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{e-color}{<^{\mu}} e_{1} & v_{8} \textcolor{closed-color}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{e-color}{<^{\mu}} e_{1} & e_{5} \textcolor{closed-color}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{e-color}{<^{\mu}} e_{1} & w_{9} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]

and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{e-color}{E}$ are depicted with a dashed box, while edges in $\textcolor{closed-color}{E}$ --- with a round solid box.
\begin{figure}
\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2_2}
}
\]
\captionsetup{skip=0pt, belowskip=-1ex}
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}
\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{e-color}{E_{\mathcal{F}}}) \subseteq \textcolor{e-color}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{closed-color}{E_{\mathcal{F}}}) \subseteq \textcolor{closed-color}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          \text{if } e \textcolor{e-color}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{e-color}{<_{\mathcal{G}}^{\mu}} \phi(v)
          \qquad
          \text{if } e \textcolor{closed-color}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          \text{if } e_1 \textcolor{e-color}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{e-color}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \qquad
          \text{if } e_1 \textcolor{closed-color}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$.
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

@@

\end{definition}
We call $n$  \textit{external input interfaces}, $n'$ \textit{internal input interfaces},
and $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  
We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n$ and $m$ with their images in $n'$ and $m'$,  and likewise $m, m, n', m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.
Graphically, we depict such extended cospans as in Fig.~\ref{fig:e-cospan-example}: feet of the cospan are depicted within shaded regions.
All non-black vertices form strictly internal interfaces.
We follow the same orientation that we chose for string diagrams --- the input interface of the cospan is at the bottom and the output interface is at the top.

### 2025-07-09 POPL2026/paper/sections/lambda-calculus.tex
figures and benchmarks

\section{Application --- $\lambda^\sim_{\text{lsub}}$-calculus}

@@

\autoref{fig:plus-1-2-programs} shows some programs which are graphically and semantically equivalent but syntactically distinct.
As alluded to in~\autoref{sec:vs-e-graphs-with-binding}, the traditional (slotted) e-graph approach to modelling the graphical equivalence of these terms (\autoref{fig:plus-1-2-terms}) relies on explicitly encoding substitution operations as e-graph nodes, and their behaviour (the graphical equivalence relation $\sim$) as rewrite rules, as in~\autoref{fig:e-graph-substitution} --- not only is this more complicated, but it also leads to a blow-up in the state space required to perform equality saturation in any extension of the $\lambda^\sim_{\text{lsub}}$-calculus, because we are really rewriting $\lambda_{\text{lsub}}$-terms and computational resources are spent rewriting between graphically equivalent terms.
\textcolor{red}{Aleksei, can you say something quantitative here if you ran some tests?}
In contrast, in our framework, $\lambda^\sim_{\text{lsub}}$-terms are directly modelled as string diagrams (\autoref{fig:plus-1-2-string-diagram}\footnote{Note that dashed boxes are elided for convenience here because no rewriting has occurred.}).
We do however note that the traditional approach is more flexible, in the sense that it may model more theories with binders (e.g.\ $\pi$-calculus), whereas our approach is essentially specialised to $\lambda$-calculus, relying heavily on the metatheoretic internal language of the ambient setting.

>>>>

\section{Application --- $\lambda^\sim_{\text{lsub}}$-calculus}
\label{sec:application}

@@

\autoref{fig:plus-1-2-programs} shows some programs which are graphically and semantically equivalent but syntactically distinct.
As alluded to in~\autoref{sec:vs-e-graphs-with-binding}, the traditional (slotted) e-graph approach to modelling the graphical equivalence of these terms (\autoref{fig:plus-1-2-terms}) relies on explicitly encoding substitution operations as e-graph nodes, and their behaviour (the graphical equivalence relation $\sim$) as rewrite rules, as in~\autoref{fig:e-graph-substitution} --- not only is this more complicated, but it also leads to a blow-up in the state space required to perform equality saturation in any extension of the $\lambda^\sim_{\text{lsub}}$-calculus, because we are really rewriting $\lambda_{\text{lsub}}$-terms and computational resources are spent rewriting between graphically equivalent terms.
\textcolor{cyan}{To illustrate this, let's consider the following equations induced by the graphical equivalence and the $\lambda_{\text{lsub}}$ reduction rules}

\begin{tabular}{lcc}
	\texttt{beta} & $(\lambda x . t) u = t[x / u]$&\\
	\texttt{subst\_same\_var} & $x[x / u] = u$ &\\
	\texttt{subst\_diff\_var} & $y[x / u] = y$ & if $x \not \in \mathcal{F}(u)$\\
	\texttt{subst\_application} & $(uu')[x / v] = u[x / v]u'[x / v]$ & \\
	\texttt{subst\_lambda} & $(\lambda x . u)[y / v] = \lambda x . (u[y / v])$ & if $x \not \in \mathcal{F}(v)$ \\
	\texttt{subst\_interchange\_1} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v[x' / v']]$ & if $x' \in \mathcal{F}(v)$ and $x \in \mathcal{F}(v')$\\
	\texttt{subst\_interchange\_2} & $(u[x / v])[x' / v'] = (u[x / v[x' / v']])$ & if $x' \not \in \mathcal{F}(u)$\\
	\texttt{subst\_comm} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v]$ & if $x \not \in \mathcal{F}(v')$ and $x' \not \in \mathcal{F}(v)~.$\\
\end{tabular}

\textcolor{cyan}{These can be encoded in a framework of slotted e-graphs.
To demonstrate the effect that a particular set of rewrites has on the number of e-nodes of an e-graph, we saturate the lambda term $plus 2 2$ where $plus$ and $2$ are encoded using Church encoding.
The numbers of e-nodes after the e-graph is saturated for each set of rewrite rules are presented in \autoref{tbl:slotted}.
}

\begin{table}
	\begin{tabular}{lcccc}
	  Set of equations & explicit subst & explicit subst & explicit subst & explicit subst\\
	  && + comm& + interchange&  + interchange\\
	  &&&&  + comm\\
	  \# of e-nodes & 161 & 288 & 289 & 356
	\end{tabular}
	\caption{Slotted e-graphs benchmarks}
	\label{tbl:slotted}
\end{table}

\textcolor{cyan}{The \texttt{explicit subst} set of rewrite rules consist of the first five rules above.
We can see that additionally saturating the e-graph with respect to interchange and commutativity rules increases the number of e-nodes by a factor of 2.
These are the exact equations that are absorbed by using string-diagrammatic representation of $\lambda$-terms.
Notably, string diagrams (and (e-)hypergraphs) do not absorb the equation \texttt{subst\_lambda} but we believe it is practically feasible to do rewriting modulo this rule by traversing through the boxes when searching for a redex.
Furthermore, (cospans of) e-hypergraphs, as presented in \autoref{sec:closed-e-hyp}, do not absorb the equations for associativity, commutativity and unitality for $\comonoid$ and $\counit$, but the constraints on the cospans and the boundary complement may be relaxed to absorb those as well by following the work of~\cite{zanassi_comonoid} with which our approach is completely compatible.}

\textcolor{red}{Aleksei, can you say something quantitative here if you ran some tests?}
In contrast, in our framework, $\lambda^\sim_{\text{lsub}}$-terms are directly modelled as string diagrams (\autoref{fig:plus-1-2-string-diagram}\footnote{Note that dashed boxes are elided for convenience here because no rewriting has occurred.}).
We do however note that the traditional approach is more flexible, in the sense that it may model more theories with binders (e.g.\ $\pi$-calculus), whereas our approach is essentially specialised to $\lambda$-calculus, relying heavily on the metatheoretic internal language of the ambient setting.

### 2025-07-09 POPL2026/paper/sections/soundness.tex
figures and benchmarks

\begin{proposition}%
	\label{prop:quotient-structural}
	For two closed $\Sigma^{+}$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
By definition, we have 
$
  f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
such that $f_{i} \leadsto f_{j}$ for some $i, j$.
This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{e-color}{E}$, we can apply Proposition~\ref{prop:fscd}.
\end{proof}

>>>>

\begin{proposition}
    For two closed $\Sigma^{+}$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
By definition, we have 
$
  f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
such that $f_{i} \leadsto f_{j}$ for some $i, j$.
This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{e-color}{E}$, we can apply Proposition~\ref{prop:fscd}.
\end{proof}
